Using Scagnostics to find Interesting Projections of Multivariate Data

### 2018-09-07 scagnostics.Rmd
Adding comments for setting up structure.

A related question is the selection of pair-wise scatter plots in datasets with large numbers of parameters, for which different methods have been developed. One approch are scatterplot diagnostics, "scagnostics", developed by @scag, and explored in @WW08. They are calculated from a set of elementary graph-theoretic building blocks: the convex hull, the alpha hull and the minimal spanning tree (MST). These are used to define eight measures: Outlying, Skewed, Sparse, Clumpy, Striated, Convex, Skinny and Stringy. While in principle each measure could be used on its own to define a ranking, the original idea suggested considering pair-wise scatter plots of the scagnostics measures for identification of "characteristic" and "unusual" 2-d views in the original parameter space. Alternative methods suggested by (Katrins Thesis or maybe her R package?) suggest using parallel coordinates, glyphs or a so-called scagramm (based on the idea of corrgrams) that shouold be coupled to interactive brushing for selection of pair-wise scatter plots to be displayed.
(FIXME just give reference rather than all definitions here?)
(Issues with scagnostics already for selection of pairs have been described in Katrins Thesis, maybe have short summary of that here?)
Given the various shortcomings of the scagnostics approach we consider several other measures that can directly be interpreted as ranking of information contained in a given 2-d view.
Two such additional measures have been defined in @mbgraphic. The first is based on smoothing splines, and is defined such that it will take maximum values when the data distribution can be described by a functional form. For a scatterplot with variable X and Y we calculate it as
\begin{equation}
splines_{2d} = max(1-\frac{Var(res_{X\sim Y})}{Var(X)}, 1-\frac{Var(res_{Y\sim X})}{Var(Y)})
\end{equation}
where $Var(.)$ is the variance and $res_{X\sim Y}$ are the residuals in the spline model where $X$ is considered a function of $Y$.
The second new measure introduced in @mbgraphic is aiming to detect also non-functional relations and is based on the idea of distance correlation @szekely2007. (should I repeat definitions here?)
Finally we will also consider the maximal and total information coefficient (MIC and TIC) defined in @Reshef1518 which are based on mutual information found when imposing grids on pair-wise variable plots. These functions are availabel in R via the interface implemented in the minerva package @minerva.

@@

A projection pursuit index, a function of all possible projections of the data, invariably has many "hills and valleys" and "knife-edge ridges" because of the varying shapes in the underlying density of observations from one projection to the next. From an exploratory data analysis perpective it is interesting to combine numerical optimisation with visualisation to watch the structure of the data moving into a maximum, to jump away from this projection and follow the optimisation again, invariably moving from local maxima to local maxima, and perhaps even a global maximum. Each of these maxima can reveal different information about the data. Projection pursuit optimisation combined with geodesic interpolations between projections provided by a tour [@As85], is called a projection pursuit guided tour [@CBCH94]. Software to run a projection pursuit guided tour is available in the R package, "tourr" [@tourr].

@@

```{r load}
library(tourr)
library(tidyverse)
library(reshape2)
library(binostics) #https://github.com/uschiLaa/binostics
library(gridExtra)
library(wesanderson) #color palette
library(tictoc) #timer
library(mbgraphic) #Katrins package
library(dfoptim)
library(GGally)
library(geozoo)
```

The nine scagnostics measures are designed to be computed together, producing a matrix of values with dimensions corresponding to the number of all pairs of variables, and the number of scagnostics. If there are 100 variables, the dimension will be $4950x9$. It is typically used to extract the pairs of variables that have the highest scagnostic values, and thus considered to be the most interesting. It can also be used to do more in-depth exploratory analysis, and is interesting to study the distribution of the scagnostics resulting from the data, as a data set in itself. In a sense this the data's "fingerprint".

>>>>

A related question is the selection of pair-wise scatter plots in datasets with large numbers of parameters, for which different methods have been developed. One approch are scatterplot diagnostics, "scagnostics", developed by @scag, and explored in @WW08. They are calculated from a set of elementary graph-theoretic building blocks: the convex hull, the alpha hull and the minimal spanning tree (MST). These are used to define eight measures: Outlying, Skewed, Sparse, Clumpy, Striated, Convex, Skinny and Stringy. While in principle each measure could be used on its own to define a ranking, the original idea suggested considering pair-wise scatter plots of the scagnostics measures for identification of "characteristic" and "unusual" 2-d views in the original parameter space. Alternative methods suggested by (Katrins Thesis or maybe her R package?) suggest using parallel coordinates, glyphs or a so-called scagramm (based on the idea of corrgrams) that shouold be coupled to interactive brushing for selection of pair-wise scatter plots to be displayed.
(FIXME just give reference rather than all definitions here?)
(Issues with scagnostics already for selection of pairs have been described in Katrins Thesis, maybe have short summary of that here?)
Given the various shortcomings of the scagnostics approach we consider several other measures that can directly be interpreted as ranking of information contained in a given 2-d view.
Two such additional measures have been defined in @mbgraphic. The first is based on smoothing splines, and is defined such that it will take maximum values when the data distribution can be described by a functional form. For a scatterplot with variable X and Y we calculate it as
\begin{equation}
splines_{2d} = max(1-\frac{Var(res_{X\sim Y})}{Var(X)}, 1-\frac{Var(res_{Y\sim X})}{Var(Y)})
\end{equation}
where $Var(.)$ is the variance and $res_{X\sim Y}$ are the residuals in the spline model where $X$ is considered a function of $Y$.
The second new measure introduced in @mbgraphic is aiming to detect also non-functional relations and is based on the idea of distance correlation @szekely2007. (should I repeat definitions here?)
Finally we will also consider the maximal and total information coefficient (MIC and TIC) defined in @Reshef1518 which are based on mutual information found when imposing grids on pair-wise variable plots. These functions are availabel in R via the interface implemented in the minerva package @minerva.
(read paper and add more detail. how are they similar/different compared to scagnostics and distance correlation? the way that binning is being built into it may make difference in how MIC/TIC may be smoother?)

@@

A projection pursuit index, a function of all possible projections of the data, invariably has many "hills and valleys" and "knife-edge ridges" because of the varying shapes in the underlying density of observations from one projection to the next. From an exploratory data analysis perpective it is interesting to combine numerical optimisation with visualisation to watch the structure of the data moving into a maximum, to jump away from this projection and follow the optimisation again, invariably moving from local maxima to local maxima, and perhaps even a global maximum. Each of these maxima can reveal different information about the data. Projection pursuit optimisation combined with geodesic interpolations between projections provided by a tour [@As85], is called a projection pursuit guided tour [@CBCH94]. Software to run a projection pursuit guided tour is available in the R package, "tourr" [@tourr].
(Even just for the optimisation part the guided tour implementation is useful.)
(Traditional projection pursuit indices are fast to compute and therefore the optimisation would be done in "real time" while displaying the tour. For the measures considered here this generally does not apply, as a consequence we "disentangle" the two into subsequent steps.)

@@

```{r load}
library(tourr)
library(tidyverse)
library(reshape2)
library(binostics) #https://github.com/uschiLaa/binostics
library(gridExtra)
library(wesanderson) #color palette
library(tictoc) #timer
library(mbgraphic) #Katrins package
library(dfoptim)
library(GGally)
library(geozoo)
library(minerva)
```

The nine scagnostics measures are designed to be computed together, producing a matrix of values with dimensions corresponding to the number of all pairs of variables, and the number of scagnostics. If there are 100 variables, the dimension will be $4950\times 9$. It is typically used to extract the pairs of variables that have the highest scagnostic values, and thus considered to be the most interesting. It can also be used to do more in-depth exploratory analysis, and is interesting to study the distribution of the scagnostics resulting from the data, as a data set in itself. In a sense this the data's "fingerprint".
